---
title: "Project 2"
subtitle: "BA 713: Data and Decisions (Waddell)"
author: "YourTeamName (member1, member2, etc)"
date: "Due: 13 March 2021 at 11pm"
output: 
 html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman") 
library(pacman)
p_load(dplyr, tidyverse, ggplot2, knitr, ggthemes, kableExtra)

# Knitr options
opts_chunk$set(
  echo=TRUE, 
  comment = "#>",
  fig.align = "center",
  fig.height = 5,
  fig.width = 8,
#  dpi = 300,
  cache = F,
  warning = F,
  message = F
)

# ggplot theme
theme_simple <- theme_bw() + 
  theme(
    plot.title = element_text(size=12),
    #axis.line = element_line(size=.5),
    #panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    rect = element_blank(),
    strip.text = element_blank(),
    axis.title = element_text(size=10),
    axis.text.x = element_text(size = 10, margin = margin(-8,0,0,0)),
    axis.text.y = element_text(size = 10, angle=90),
    axis.ticks = element_blank()
  )
```

# Intro

In the code chunk below, you'll see that I have read in some election data, and plotted `newdeath.pc` (from Project 1) for every state, for all days in 2021 (through Feb 27), faceted by whether the state went for the Democrat or Republican candidate. I've also used a gradient to indicate the vote share. (_It's a mess._)

Note the use of `stringr::str_to_title()`, `filter()`, `lm()`, and `huxtable::huxreg()`. They're new to you. 

```{r, echo=T}
covid <- read.csv("https://covidtracking.com/data/download/all-states-history.csv")
covid <- covid %>% select(state, date, death, hospitalized)
covid <- covid[is.na(covid$death)==F, ]  # drop missing deaths
covid <- covid %>%
    group_by(state) %>% 
    mutate(
      newdeath = death - lead(death),
      newhospitalized = hospitalized - lead(hospitalized)
    ) %>% ungroup()
covid <- covid[is.na(covid$newdeath)==F, ]  # drop missing newdeaths
covid <- covid %>%
  mutate(
    newdeath = ifelse( newdeath<0, 0, newdeath),
    newhospitalized = ifelse( newhospitalized<0, 0, newhospitalized),
    NAME = state.name[match(state, state.abb)],
    date = as.Date(date),
    month = months(date),
    year = format(date, format="%Y")
  )
covid$date <- as.Date(covid$date)

# Census populations
census.pop <- read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/national/totals/nst-est2020.csv") %>%
  select( NAME, POPESTIMATE2020) 

# election data
p_load(stringr)
election <- read.csv("https://www.dropbox.com/s/8uvoxi4vjb7kzll/1976-2020-president.csv?dl=1") %>%
  filter(year==2020 & (party_detailed=="DEMOCRAT" | party_detailed=="REPUBLICAN")) %>%
  select(state, party_detailed, candidatevotes) %>%
  group_by(state) %>%
  mutate(
    totalvotes.DRonly = sum(candidatevotes),
    share.D = 100*ifelse(party_detailed=="DEMOCRAT", 
                    candidatevotes / sum(candidatevotes), 
                    1 - (candidatevotes / sum(candidatevotes))),
    share.R = 100*ifelse(party_detailed=="REPUBLICAN", 
                    candidatevotes / sum(candidatevotes), 
                    1 - (candidatevotes / sum(candidatevotes)))
  ) %>% ungroup() %>%
  filter(party_detailed=="REPUBLICAN") %>% 
  select(state, share.R, share.D, totalvotes.DRonly)

election$state = stringr::str_to_title(election$state)

# merging 
df <- merge( covid, census.pop, by="NAME")
df <- merge( df, election, by.x="NAME", by.y="state")
df <- df %>% 
  mutate(
    newdeath.pc = 1e6*(newdeath/POPESTIMATE2020),
    death.pc = 1e6*(death/POPESTIMATE2020),
    totalvotes.DRonly.pc = totalvotes.DRonly/POPESTIMATE2020,
    winning_party = ifelse(share.D>50,"Democrat states","Republican states"),
    POPESTIMATE2020.mil = POPESTIMATE2020/1e6
  ) %>% 
  filter(date>="2021-01-01")

# note oddities
#summary(df$newdeath.pc)  # a couple outliers
gg <- df %>% 
  filter(
    newdeath.pc<218
  )
ggplot(data=gg, 
       aes(x=date, y=newdeath.pc, group=factor(state), color=share.R)) +
  geom_line(alpha=.7, size=.1) + 
  labs(
    title = "Daily Covid-19 Deaths",
    y = "Deaths per million (daily)",
    x = "\nDate",
    caption = "Sources: Covid-19 data from Covid Tracking Project; Election data from MIT Election Data + Science Lab",
    color = "Republican share of votes"
  ) +
  theme_simple +
  theme(
    plot.title = element_text(size=16),
    axis.text.x = element_text(vjust=0.5,angle=90),
    strip.text.x = element_text(size = 12)
  ) + 
  scale_color_gradient(low="blue", high="red") +
  scale_x_date(date_labels = "%b-%Y") 
  # facet_wrap(~ winning_party)  #THIS LINE WAS CAUSING ERRORS ON SOME CLOUD SERVERS, SO I'VE COMMENTED IT OUT FOR NOW
```


# Part 1

Below, I've written code to run two models. In the first, I simply ask if a state's deaths move with the republican vote share. In the second, I control for the share of the population that voted. (_In each, I also control for a cubic trend. I don't show this in the table so we can focus a bit._)

``` {r, echo=T}
df.reg <- df %>%
  filter(date>="2021-01-01" & date<="2021-02-27") %>%
  mutate(t = as.numeric(date-min(date)),
         t2 = t*t,
         t3 = t*t2
    )

m1 <- lm(newdeath.pc ~ share.R + t + t2 + t3, df.reg)
m2 <- lm(newdeath.pc ~ share.R + totalvotes.DRonly.pc + t + t2 + t3, df.reg)

p_load(huxtable)
ht <- huxreg("Model 1" = m1, "Model 2" = m2,
             statistics = character(0),
             coefs = c("Republican share of votes (0 to 100)" = "share.R", 
                       "Total votes / State population (0 to 1)" = "totalvotes.DRonly.pc"),
             note = "{stars}. t statistics in parentheses."
       )
width(ht) <- .75
wrap(ht) <- FALSE
ht %>% set_caption("Daily Covid-19 Deaths, 2021-01-02 through 2021-02-27")
```

**ACTIONABLE:** Interpret the above regression output. 

- _In Model 1, what does the_ `r round(m1$coefficients[2],3)` _mean?_ Answer here...

- _In Model 1, what do the stars on_ `r round(m1$coefficients[2],3)` _mean?_ Answer here...

- _How do you imagine Model 1 being used to infer something about Republican-leaning states and Covid deaths?_ Answer here...

- _Would it be reasonable to interpret Model 1 that way? Why? Why not?_ Answer here...

- _What must be true for the relationship between Covid deaths and "Republican share of votes" to change between Model 1 and Model 2?_ Answer here...

- _In Model 2, what does the_ `r round(m2$coefficients[2],3)` _mean?_ Answer here...

- _In Model 2, what does the lack of stars on_ `r round(m2$coefficients[2],3)` _mean?_ Answer here...


<br>
---
<br>


# Part 2

Here, I am most interested in the consistency that follows from your research question to your methodology&mdash;be careful about the question, and be sure that the methods follow.  


Picking up where we left off with Part 4 of Project 1...


**ACTIONABLE 1** What is the causal question you are asking?

- respond here

**ACTIONABLE 2** Why do you care about knowing about this relationship? Why should anyone else care?

- respond here 

**ACTIONABLE 3** What would the ideal experiment be were you to be able to conduct the ideal experiment?

- respond here

**ACTIONABLE 4** How will you approach answering this question, and how does it approximate the ideal experiment? (_What must you assume is random in order to interpret the results as causal?_)

- respond here

**ACTIONABLE 5** Include your analysis in the chunk below.

``` {r, echo=T}
# respond here

```


---
<br><br><br><br>



